---
title: "Applied Finance: European Up-and-In Put Option"
subtitle: "Path-dependent option pricing with Monte Carlo and the `Rcpp` package"
author: "Dustin Pacholleck - 437968"
date: "Academic Year 2022/2023"
output: 
  html_document:
    toc: true
    theme: journal
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# Objective

This project aims to use Monte Carlo Simulation for pricing a European style up-and-in put option with a barrier active between the moment of pricing and the option expiry.

Specifically, the project tries to:

1. find the theoretical price of the given option,
2. show the relation between the theoretical price of the given option and two factors (simultaneously):
    a) volatility of the underlying instrument returns, and
    b) time to maturity of the option.

It tries to achieve so, by building a `rccp` package. 

# Assumptions

The project assumes: 

* 10,000 Monte Carlo simulations,
* 252 number of time steps per year, and
* defining the barrier after inspection of its effect.


# Description of the Option

The given option is described as followed: 

* price of the underlying at the moment of option pricing: $S0 = 95$,
* strike price $K = 100$,
* annualized volatility rate $\sigma = 0.24$ 
* annualized risk-free rate $r = 0.07$
* time to maturity $t = 0.75$
  
A holder of an European option can exercise only at exercise date. The option includes an Up-and-In barrier which means the underlying price has to be higher than the barrier to activate the option.

In finance, an **option** is an agreement, which gives the holder a right for **buy** or **sell** a certain underlying instrument for a given price (*exercise price*) at a given moment in futures (*expiry date*).
**PUT** option -- gives the holder right to **sell** a given amount of underlying instrument for a given price at the given moment in future. 
`(Source: Lecture Slides)`

# Interpretation of the results

```{r Setup, message=FALSE, warning=FALSE, include=FALSE}

## Installing Packages
# install.packages("devtools")
# install.packages('tidyverse')
# install.packages('here')

## Loading packages
library(devtools)
library(tidyverse)
library(here)

setwd(here()) # adjust if needed

###### Options to install the option pricer package ###### 
## Please make sure it is not currently installed:
#
# remove.packages("optionPricer")

## 1) install package from file
# 
# install.packages("Packages/optionPricer_1.0.tar.gz",
#                  type = "source",
#                  repos = NULL)

## 2) install package from GitRepo
# source('secret.R') # my Git Key is stored in a separate file
# auth_token <- 'Please add your auth token here (https://github.com/settings/tokens)'
# devtools::install_github('Pacholleck/OptionPricing/Packages/optionPricer',auth_token=auth_token)

library(optionPricer)

###################################################################################################

#' Calculate the theoretical price of a European style up-and-in put option using Monte Carlo simulation
#' @param nInt the number of time steps per year
#' @param strike the strike price of the option
#' @param spot the initial price of the underlying instrument
#' @param vol the volatility of the underlying instrument returns
#' @param r the risk-free interest rate
#' @param expiry the time to maturity of the option (in years)
#' @param B the barrier level
#' @param nReps the number of Monte Carlo simulations to run
#' @return the approximated theoretical price of the up-and-in put option
#' @example
#' getEuropeanPutPrice_R(126,100,95,0.24,0.07,0.75,125,1000)

## Setting nReps as default tp 10.000 for easy computation. Please adjust if wanted.
nReps <- 10000

```

## Determining the Barrier
```{r Barrier, echo=FALSE, message=FALSE, warning=FALSE}


# R wrapping function: general
getEuropeanPutPrice_withB <- function(B) {
  return(optionPricer::getEuropeanPutPrice(252,100,95,0.24,0.07,0.75,B,nReps))
}

# arguments values of values of function 
B <- seq(50,250, by = 5)
prices <- sapply(B, getEuropeanPutPrice_withB)

# visualization: options price vs. expiry 
tibble( B, prices) %>%
  ggplot(aes(B, prices)) +
  geom_point(col = "red") +
  labs(
    x     = "Barrier",
    y     = "Option Price",
    title = "Price of European Up-and-In Put Option vs. Barrier",
    caption = "source: own calculations with the optionPricer package")
```

The barrier seems to have an effect within the range of 90 to 160 on the option price. Therefore, for this project we will assume the barrier to be $B = 125$.

## Calculating the Theoretical Price 
```{r Price Estimation, echo=FALSE, message=FALSE, warning=FALSE}

# R wrapping function: general
getEuropeanPutPrice_R <- function(nInt, strike, spot, vol, r, expiry, B, nReps) {
  return(optionPricer::getEuropeanPutPrice(nInt, strike, spot, vol, r, expiry, B, nReps))
}

# Estimate theoretical price of the given option
price <- getEuropeanPutPrice_R(252,100,95,0.24,0.07,0.75,125,nReps)

```

The theoretical option price is estimated to be **`r price`** for the given option. 

## Investigating the Relationship of Price, Volatility and Time to Maturity 
```{r Relationship, echo=FALSE, message=FALSE, warning=FALSE}

# R wrapping function: option price vs. volatility and maturity
getEuropeanPutPrice_WithVolAndExpiry <- function (vol, expiry) {
  return(
    optionPricer::getEuropeanPutPrice(252,100,95,vol,0.07,expiry,125,nReps))
}

# sequences of argument values
vol <- seq(0, 1, by = 0.1)
expiry  <- seq(0, 1, by = 0.1)
# setup grid with arguments
grid      <- expand.grid(vol = vol, expiry = expiry)
# get prices for grid
prices    <- mapply(getEuropeanPutPrice_WithVolAndExpiry, 
                    vol = grid$vol, expiry = grid$expiry)
result.df <- data.frame(grid, prices)

# visualization: option price vs. volatility and maturity
grid %>% 
  as_tibble() %>%
  bind_cols(price = prices) %>%
  ggplot(aes(x = vol, y = price, group = as.factor(expiry), colour = as.factor(expiry))) +
  geom_line() +
  geom_point(size = 1, shape = 21, fill = "white") +
  labs(
    x     = "Volatility",
    y     = "Option Price",
    title = "Price of European Up-and-In Put Option vs. Volatility and Time to Maturity",
    caption = "source: own calculations with the optionPricer package",
    colour='Time to Maturity')

```

1. **time to maturity** -- influences positively the put option values. The longer the time to maturity, the higher the probability that the option becomes active and in-the-money.

2. **volatility of underlying prices** -- influences positively the put option values. The higher the volatility, the higher the probability that the option becomes active and in-the-money.

# Code of Honor
In accordance with the Honor Code, I certify that my answers here are my own work, and I
did not make my solutions available to anyone else.
